---
title: "Uncertainty"
author: Cory Whitney, Eike Luedeling 
output:
      rmarkdown::html_vignette
bibliography:
 - bib/packages.bib
 - bib/papers.bib
vignette: >
  %\VignetteIndexEntry{Uncertainty}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Print pdf and word versions -->
<!-- rmarkdown::render("uncertainty-vignette.Rmd", output_format = "pdf_document") -->
<!-- rmarkdown::render("uncertainty-vignette.Rmd", output_format = "word_document") -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, warning = F, include = F}
#install and load packages
library(ggplot2)
library(rmarkdown)
# devtools::install_github("CWWhitney/uncertainty")
library(uncertainty)

#Automatically write R package citation entries to a .bib file
knitr::write_bib(c(.packages(), 
                   'ggplot2'), 'bib/packages.bib')

#resolves an issue with the scientific style of numbers, like one can read inside this answer from @Paul Hiemstra: https://stackoverflow.com/a/25947542/4061993 from the documentation of ?options:
options(scipen=999)
```

Create random influence variable {in_var} and outcome variable {out_var} data using `sample()`. 

```{r}
in_var <- sample(x = 10:500, size = 20, replace = TRUE)
out_var <- sample(x = 1000:5000, size = 20, replace = TRUE)
```

Using `varscatter()` create a scatter plot of an influence variable of interest {in_var} (x) and an outcome variable {out_var} (y) with associated and estimated densities with loess smooth linear fits density curves. Plot made with the `scatter.hist()` function in the `psych()` package [@R-psych] in R [@R-base].  

```{r, fig.width=7, fig.height=7}
uncertainty::varscatter(in_var, out_var)
```

Use `varkernel()` for another type of graphical representation of the same relationship. `varkernel()` performs a two-dimensional kernel density estimation for the expected value of the outcome variable {out_var} and the influence variable of interest {in_var} using the `kde2d()` function in the `MASS()` package from base R [@R-base]. The function produces a matrix of the estimated density (z) of the expected value of the outcome variable {out_var} (y) and the influence variable of interest {in_var} (x). It looks slightly different than the scatter plot estimates above because the density function restricts the shape of the kernel to a bivariate normal kernel.

```{r, fig.width=7, fig.height=7}
uncertainty::varkernel(in_var, out_var)
```

`varkernel()` shows a density surface plot of a given influence variable of interest {in_var} (x) and the outcome variable {out_var} (y). The legend shows the value for the estimated density (z). The plot is made with the `filled.contour()` function of the `graphics()` package.

In `varkernel()` the density (z) over the entire plot integrates to one, and therefore represents the relative probability of an observation (the outcome variable {out_var} along y-axis) given a specific value for a given influence variable of interest {in_var} (along x-axis). 

## Estimated outcome variable {out_var} given the expected variable

`varkernelslice()` calculates the estimated outcome variable {out_var} given the expected value of the influence variable of interest {in_var}, based on a slice of 'z' from the Kernel density calculated with `varkernel()`. The `expectedvariable` parameter is set to 30.

```{r, fig.width=7, fig.height=7}
uncertainty::varkernelslice(in_var, out_var, expectedin_var = 30)
```

`varkernelslice()` plots the probabilities (shown along the y-axis) for the expected value of the outcome variable {out_var} (shown along the x-axis). Since this is a cut through the density kernel `varkernel()`, which integrates to 1, the probability values are relative, not absolute measures.

## Influencing variable intervals

`varviolin()` determines different possible influencing variable intervals {in_var} by calculating the optimal interval width for the given variable values. This is done using the `IQR()` function in the `stats()` package, after the Freedman-Diaconis rule (IQR = interquartile range).
`Optimal interval width for our sample = 2 * interquartile range for our sample / (total number of observations for the interquartile range for our sample)^(1/3)`

```{r, fig.width=7, fig.height=7}
uncertainty::varviolin(in_var, out_var)
```

`varviolin()` shows violin plots of influencing variable {in_var} values (x) and outcome variable {out_var} (y) values, with six different intervals of influencing variable {in_var} values. Plot made with `ggplot2()` [@R-ggplot2].

## Probability of outcome variable given variable

`varkernelslicerange()` shows the influencing variable {in_var} value intervals, optimized interquartile ranges shown in `varviolin()` can be used to select a range to slice from the density kernel `varkernel()` as was done for a single variable value in `varkernelslice()`. Here we set the maximum variable value to 200 and the minimum to 100.

```{r, fig.width=7, fig.height=7}
uncertainty::varkernelslicerange(in_var = in_var, out_var = out_var, 
                                 min_in_var = 100, max_in_var = 200)
```

`varkernelslicerange()` plots the probabilities (shown along the y-axis) for the expected value of the outcome variable {out_var} (shown along the x-axis). As with `varkernelslice()` the probability values shown are relative, not absolute measures. They are the result of cuts through the density kernel `varkernel()`, which integrates to 1. 

## Difference in probability of outcome variable given variable

To tell the difference in two distributions determined by the `varkernelslice()` ranges we can test the difference. We can extract the data with the `varkernelslice_data()` function.

```{r}

A <- varkernelslice_data(in_var, out_var, expectedin_var = 40)
B <- varkernelslice_data(in_var, out_var, expectedin_var = 10)

sum(A < sample(B, length(A))) / length(A)

```

This document was generated using R Markdown [@R-rmarkdown].

# References
